The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2020-07-13 journal paper/content/sec-intro.tex
updates

\xhdr{Economic interpretation: network effects of data.}
Our model speaks to the ongoing debate on the role that data plays in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional ``network effects", \asedit{creating scenarios when only one firm can function in the market \citep{Rysman09, jullien2019economics}.}
%whereby, when these effects are present, in many cases only one firm can function in the market, leading to competition \emph{for} the market being more important than competition \emph{in} the market \citep{Rysman09, jullien2019economics}.
\asedit{The death spiral/amplification effects mentioned above have a similar flavor as network effects: a relatively small amount of exploration (resp., data advantage)  gets amplified under competition and causes the firm to be starved out of users (resp., take over most of the market).}
%\gaedit{We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market, showing that data can provide a similar incumbency advantage as those provided by traditional network effects and can serve as a barrier to entry in online markets.}
A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model the network effects, but they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \citet{lambrecht2015can, bajari2018impact} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.

>>>>

\xhdr{Economic interpretation: network effects of data.}
Our model speaks to the ongoing debate on the role that data plays in the digital economy, \gaedit{which has been at the forefront of policy discussions on how to regulate major, data-intensive, digital platforms \citep{competition2019digital, scott2019committee}.} One fundamental question in this debate is whether data can serve a similar role as traditional ``network effects", \asedit{creating scenarios when only one firm can function in the market \citep{Rysman09, jullien2019economics}.}
%whereby, when these effects are present, in many cases only one firm can function in the market, leading to competition \emph{for} the market being more important than competition \emph{in} the market \citep{Rysman09, jullien2019economics}.
\asedit{The death spiral/amplification effects mentioned above have a similar flavor as network effects: a relatively small amount of exploration (resp., data advantage)  gets amplified under competition and causes the firm to be starved out of users (resp., take over most of the market).}
%\gaedit{We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market, showing that data can provide a similar incumbency advantage as those provided by traditional network effects and can serve as a barrier to entry in online markets.}
A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model the network effects, but they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.

### 2020-07-13 journal paper/ec19paper/content/perf_in_iso.tex
updates

\begin{figure}
\centering
\includegraphics[scale=0.35]{ec19paper/figures/nih_iso_mean}
\includegraphics[scale=0.35]{ec19paper/figures/uniform_mean}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\caption{Mean reputation trajectories for Needle-in-Haystack and Uniform. The shaded area shows 95\% confidence intervals.}
\label{prelim_means}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[scale=0.35]{ec19paper/figures/nih_mean}
\includegraphics[scale=0.35]{ec19paper/figures/mean_inst_reward_nih}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\caption{Mean reputation trajectory (left) and mean instantaneous reward trajectory (right) for Needle-in-Haystack. The shaded area shows 95\% confidence intervals.}
\label{prelim_means}
\end{figure}
